---
title: "Milestone 4 and 5"
author: "Sandya Krishna, Mai Sedki, Lourdes Valdez"
date: "11/30/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Research questions / Problem statement:
Infectious diseases are a very important public health issue. So we want to 
examine overall communicable disease rates and trends over time of Infectious diseases reported in california. Sexually transmitted diseases will be analized separately from other groups of infectious diseases.

# Datasets to be used :
1.Infectious Diseases by County, Year and Sex (in California)2001-2018
   Source : https://data.chhs.ca.gov/dataset/infectious-disease
   Raw format of dataset:
  https://data.chhs.ca.gov/dataset/03e61434-7db8-4a53-a3e2-1d4d36d6848d/resource
  /75019f89-b349-4d5e-825d-8b5960fc028c/download/idb_odp_2001-2018.csv
  Name/source: CHHS Open Data
  Number of columns: 9
  Number of rows: 154,344
  Timing: The years included in this dataset is 2001 to 2018
  
2.STD's in California by disease, county, year and sex. Dataset: case counts
  and rates for sexually transmitted diseases (chlamydia, gonorrhea, and all
  forms of syphilis) reported for California residents.
  https://data.chhs.ca.gov/dataset/stds-in-california-by-disease-county-year-and
  -sex
  
  Name/Source: CHHS Open Data
  Number of Columns: 10
  Number of Rows: 9,558
  Timing: The years included in this dataset is 2001 to 2018
  
```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#Loading libraries:
library(tidyverse)
library(readr)
library(knitr)
library(kableExtra)
library(plotly)
library(ggplot2)
library(DT)
library(formattable)
```


```{r, include =FALSE}
#Calling the dataset and naming dataframe id_data:
id_data <- read.csv("idb_odp_2001-2018 (1) (1).csv")
```


```{r, echo=FALSE}
# Now will look at the dataframe from the environment plus the first five rows
# using the head function. Noticed the following:
#The "rate" column has a lot of "dashes" or missing values because of 0 cases
#forsome diseases. Will look at all the values to see using the function unique
#unique(id_data$rate), and other than a dash, we see empty cells and "SC"
# Will now clean the data by eliminating empty values ," ", "-" or NA 
# For correcting the NA values I will have to correct at the top of this code
# when I call the dataframe. I will do that now by adding to the code: 
#na = (c(" ","-","SC","NA"))
```
```{r include=FALSE}
id_data <- read_csv("idb_odp_2001-2018 (1) (1).csv", na = (c(" ","-","SC","NA")))
```
```{r, include=FALSE}
# Change the column names to be more r friendly,using the function rename  column names will changeto all lower case and snake_case.
id_data <- rename_with(id_data, ~ tolower(gsub(" ","_",.x,fixed=TRUE)))
```
```{r, include =FALSE}
head(id_data)
```


```{r, include=FALSE}
# next We would like to do the following changes
# to the rate column: change it to numeric, round it to 2 decimal values and d/c 
# the asterisk "*" 
id_data$rate <- as.numeric(as.character(id_data$rate))
# I view the table and run str() and now rate is numeric and asterisk is gone.
# Now will round to 2 decimals, will also round the CI to 2 decimals
id_data$rate <- round(id_data$rate,2)
id_data$`lower_95%_ci` <- round(id_data$`lower_95%_ci`, 2)
id_data$`upper_95%_ci` <- round(id_data$`upper_95%_ci`,2)

```


\newpage
# Creating variables for data analysis:
*We created new groups of variables to facilitate data presentation and analysis.
The new groups of variables are:  
1. Name of california region, for the 10 different California regions.

2. Type of infectious disease : to group each of thereported diseases by "type
of disease" , following conventional microbiology classification.

3. We also grouped years in groups of 3.

# California regions:
Superior <- "NEVADA","PLACER","PLUMAS","SACRAMENTO","SHASTA","SIERRA",
"SISKIYOU","SUTTER","TEHAMA", "YOLO", "YUBA", "MODOC", "EL DORADO", "BUTTE",
"GLENN", "LASSEN"
North Coast <- "DEL NORTE", "HUMBOLDT", "LAKE", "MENDOCINO", "NAPA","SONOMA",
"TRINITY"
Bay area<- "ALAMEDA","CONTRA COSTA", "MARIN", "SAN FRANCISCO", "SAN MATEO", 
             "SANTA CLARA", "SOLANO"
North San Joaquin Valley <- "ALPINE", "AMADOR", "CALAVERAS", "MADERA","MARIPOSA",
"MERCED", "MONO","SAN JOAQUIN", "STANISLAUS", "TUOLUMNE" 
Central Coast <- "MONTEREY", "SAN BENITO", "SAN LUIS OBISPO", "SANTA BARBARA",
             "SANTA CRUZ", "VENTURA"
South San Joaquin Valley <- "FRESNO","INYO", "KERN", "KINGS", "TULARE"
Inland Empire<- "RIVERSIDE", "SAN BERNARDINO"
LA County <- "LOS ANGELES"
Orange County <- "ORANGE"
San Diego and Imperial County <- "IMPERIAL", "SAN DIEGO"
We will also have "California" as a total for the State.

#Groups of infectious diseases:
1. Parasitic <- c("Amebiasis","Babesiosis", "Cryptosporidiosis", "Cyclosporiasis",
"Cysticercosis or Taeniasis", "Malaria", "Giardiasis", "Trichinosis")
2. Toxin_related <- c("Botulism, Foodborne","Botulism, Other", "Botulism, Wound",
"Ciguatera Fish Poisoning", "Domoic Acid Poisoning","Paralytic Shellfish
Poisoning", "Scombroid Fish Poisoning") 
3. viral <- c("Chikungunya Virus Infection", "Dengue Virus Infection","Flavivirus
Infection of Undetermined Species","Hantavirus Infection","Hepatitis E acute
infection","Rabies, human","Yellow Fever", "Zika Virus Infection") 
prions <- c("Creutzfeldt-Jakob Disease and other Transmissible Spongiform
Encephalopathies")
4. fungal <- c("Coccidioidomycosis")
5. Bacterial <- c("Anaplasmosis", "Anaplasmosis and Ehrlichiosis", "Anthrax",
"Brucellosis", "Campylobacteriosis","Cholera","E. coli O157","E. coli Other STEC
(non-O157)", "Legionellosis","Leprosy (Hansen's Disease)", "Leptospirosis",
"Listeriosis", "Lyme Disease","Plague, human","Q Fever","Spotted Fever
Rickettsiosis", "Streptococcal Infection (cases in food and dairy workers)",
"Ehrlichiosis", "Psittacosis", "Salmonellosis", "Shigellosis", "Tularemia",
"Typhoid Fever", "Paratyphoid Fever", "Typhus Fever", "Relapsing Fever", "Shiga
toxin-producing E. coli (STEC) without Hemolytic Uremic Syndrome (HUS)",
"Vibrio Infection (non-Cholera)", "Shiga Toxin Positive Feces (without culture
confirmation)","Yersiniosis")
6. Infectious_complications <- c("Hemolytic Uremic Syndrome (HUS) without 
evidence of Shiga toxin-producing E. coli (STEC)","Hemolytic Uremic Syndrome(HUS)", 
"Shiga toxin-producing E. coli (STEC) with Hemolytic Uremic Syndrome (HUS)")

# Years groups:
"2001-2003", "2004-2006", "2007-2009", "2010-2012", "2013-2015", "2016-2018"

\newpage

```{r include=F}
#Creating the first variable "region"
#We will group the counties that correspond to the California regions, as done
#by the Census bureau: For census purposes, California is divided in 10 
#regions https://census.ca.gov/regions/
#First I will use  function unique() to see all the values

#Next  will create vectors with function c() to group into the appropiate region
#looking carefully on how counties are spelled
region1 <- c("NEVADA","PLACER","PLUMAS","SACRAMENTO","SHASTA","SIERRA",
"SISKIYOU","SUTTER","TEHAMA", "YOLO", "YUBA", "MODOC", "EL DORADO", "BUTTE",
"GLENN", "LASSEN")
region2 <- c("DEL NORTE", "HUMBOLDT", "LAKE", "MENDOCINO", "NAPA","SONOMA",
"TRINITY")
region3 <- c("ALAMEDA","CONTRA COSTA", "MARIN", "SAN FRANCISCO", "SAN MATEO", 
             "SANTA CLARA", "SOLANO")
region4 <- c("ALPINE", "AMADOR", "CALAVERAS", "MADERA", "MARIPOSA",
"MERCED", "MONO","SAN JOAQUIN", "STANISLAUS", "TUOLUMNE") 
region5 <- c("MONTEREY", "SAN BENITO", "SAN LUIS OBISPO", "SANTA BARBARA",
             "SANTA CRUZ", "VENTURA")
region6 <- c("FRESNO","INYO", "KERN", "KINGS", "TULARE")
region7 <- c("RIVERSIDE", "SAN BERNARDINO")
region8 <- c("LOS ANGELES")
region9 <- c("ORANGE")
region10 <- c("IMPERIAL", "SAN DIEGO")
California <- c("CALIFORNIA")

```
```{r, include=F}
#Now will create new column using function mutate and add to current table,
I #noticed aside for counties there is a "CALIFORNIA" value, will add that
IDtable1 <- id_data %>% mutate(region = case_when(
  county %in% region1 ~ "Superior",
  county %in% region2 ~ "N_coast",
  county %in% region3 ~ "Bay_area",
  county %in% region4 ~ "NSJ_valley",
  county %in% region5 ~ "C_coast",
  county %in% region6 ~ "SSJ_valley",
  county %in% region7 ~ "I_Empire",
  county %in% region8 ~ "LA_county",
  county %in% region9 ~ "O_county",
  county %in% region10 ~ "SDI_county",
  county %in% California ~ "California",
  TRUE ~ NA_character_
)) %>%
  drop_na(region)

#it worked new column formed  called "group" on ID_table1
```
```{r, include = FALSE}
# Creating the second variable "ID_type"
# Our group decided to create the seven following groups:
parasitic <- c("Amebiasis","Babesiosis", "Cryptosporidiosis", "Cyclosporiasis",
"Cysticercosis or Taeniasis", "Malaria", "Giardiasis", "Trichinosis")
toxin_related <- c("Botulism, Foodborne","Botulism, Other", "Botulism, Wound",
"Ciguatera Fish Poisoning", "Domoic Acid Poisoning","Paralytic Shellfish
Poisoning", "Scombroid Fish Poisoning") 
viral <- c("Chikungunya Virus Infection", "Dengue Virus Infection","Flavivirus
Infection of Undetermined Species","Hantavirus Infection","Hepatitis E acute
infection","Rabies, human","Yellow Fever", "Zika Virus Infection") 
prions <- c("Creutzfeldt-Jakob Disease and other Transmissible Spongiform
Encephalopathies")
fungal <- c("Coccidioidomycosis")
bacterial <- c("Anaplasmosis", "Anaplasmosis and Ehrlichiosis", "Anthrax",
"Brucellosis", "Campylobacteriosis","Cholera","E. coli O157","E. coli Other STEC
(non-O157)", "Legionellosis","Leprosy (Hansen's Disease)", "Leptospirosis",
"Listeriosis", "Lyme Disease","Plague, human","Q Fever","Spotted Fever
Rickettsiosis", "Streptococcal Infection (cases in food and dairy workers)",
"Ehrlichiosis", "Psittacosis", "Salmonellosis", "Shigellosis", "Tularemia",
"Typhoid Fever", "Paratyphoid Fever", "Typhus Fever", "Relapsing Fever", "Shiga
toxin-producing E. coli (STEC) without Hemolytic Uremic Syndrome (HUS)",
"Vibrio Infection (non-Cholera)", "Shiga Toxin Positive Feces (without culture
confirmation)","Yersiniosis")
infectious_complications <- c("Hemolytic Uremic Syndrome (HUS) without evidence
of Shiga toxin-producing E. coli (STEC)","Hemolytic Uremic Syndrome (HUS)", 
"Shiga toxin-producing E. coli (STEC) with Hemolytic Uremic Syndrome (HUS)")

# now I will create a new table naming it IDtable2 with new column ID_type
IDtable2 <- IDtable1 %>% mutate(ID_type = case_when(
  disease %in% parasitic ~ "Parasitic",
  disease %in% toxin_related ~ "Toxin_related",
  disease %in% viral ~ "Viral",
  disease %in% prions ~ "Prions",
  disease %in% fungal ~ "Fungal",
  disease %in% bacterial ~ "Bacterial",
  disease %in% infectious_complications ~ "ID_complication",
  TRUE ~ NA_character_
)) %>%
  drop_na(ID_type)

```
```{r,include=F}
#Create ID_tableyears_grouptotal . Grouping every 3 years as a single period and change years to character, keep only total instead of male and female.
ID_tableyears_group_total <- IDtable2 %>%
  mutate(time_period =case_when(
    year %in% c(2001,2002,2003)~ "2001-2003",
    year %in% c(2004,2005,2006)~ "2004-2006",
    year %in% c(2007,2008,2009)~ "2007-2009",
    year %in% c(2010,2011,2012)~ "2010-2012",
    year %in% c(2013,2014,2015)~ "2013-2015",
    year %in% c(2016,2017,2018)~ "2016-2018",
     )) %>%
  filter(sex=="TOTAL")

```
```{r, include=FALSE}
IDtable3 <- IDtable2 %>%
  filter(sex=="TOTAL",county=="CALIFORNIA")%>%
  mutate(rate_disease= (cases/population)*100000)
```


```{r , include=FALSE}

populationv <-c(34512742, 34938290, 35388928, 35752765, 35985582, 36246822, 36552529, 36856222, 37077204, 37334578, 37678534, 38045271, 38425695, 38756940, 39076128,39328337, 39610556,39825181,34512742, 34938290, 35388928, 35752765, 35985582, 36246822, 36552529, 36856222, 37077204, 37334578, 37678534, 38045271, 38425695, 38756940, 39076128,39328337, 39610556,39825181, 34512742, 34938290, 35388928, 35752765, 35985582, 36246822, 36552529, 36856222, 37077204, 37334578, 37678534, 38045271, 38425695, 38756940, 39076128,39328337, 39610556,39825181,34512742, 34938290, 35388928, 35752765, 35985582, 36246822, 36552529, 36856222, 37077204, 37334578, 37678534, 38045271, 38425695, 38756940, 39076128,39328337, 39610556,39825181)


ID_table3californiaonly <- IDtable2 %>%
    filter(sex=="TOTAL", county=="CALIFORNIA")%>%
 filter(ID_type=="Bacterial"|ID_type== "Parasitic"|ID_type=="Fungal"|ID_type=="Viral")%>%
 group_by(ID_type,year)%>%
    summarize(sum_cases=sum(cases))%>%
  cbind(populationv)%>%
  rename(pop_total="...4")%>%
 mutate(rate_total= (sum_cases/pop_total)*100000)%>%
mutate(time_period =case_when(
    year %in% c(2001,2002,2003)~ "2001-2003",
    year %in% c(2004,2005,2006)~ "2004-2006",
    year %in% c(2007,2008,2009)~ "2007-2009",
    year %in% c(2010,2011,2012)~ "2010-2012",
    year %in% c(2013,2014,2015)~ "2013-2015",
    year %in% c(2016,2017,2018)~ "2016-2018",
     ))

#ID_table3californiaonly <- rep_len(populationv,length.out = 72)
```


\newpage

# Tables and codes
```{r}

# Tables

my_table_data <- ID_tableyears_group_total %>%
  select(c("ID_type","region","rate","time_period")) %>%
  filter(ID_type=="Bacterial"|ID_type== "Parasitic"|ID_type=="Fungal"|ID_type=="Viral") %>%
  filter(region=="California")%>%
  drop_na(rate) %>%
  group_by(ID_type,time_period,region) %>%
  summarise(cumm_rate = sum (rate))
```



```{r, include=F}

my_new_table_data <- my_table_data %>%
  pivot_wider(names_from=c(ID_type),values_from= "cumm_rate")
```


```{r}
kable(my_new_table_data, 
      booktabs=T, 
      col.names=c("Time Period", " ","Bacterial", "Fungal", "Parasitic", "Viral"),  
      align='lccc', 
      caption="Infectious disease rates (Cases/100,000) over time by disease etiology (from 2001 - 2018 by 3 year increments)",
      format.args=list(big.mark=","))
```
 

```{r, include=F}
##Mai
my_bayarea_table <- ID_tableyears_group_total %>%
  select(c("ID_type","region","rate","time_period")) %>%
  filter(ID_type=="Bacterial"|ID_type== "Parasitic"|ID_type=="Fungal"|ID_type=="Viral") %>%
  filter(region=="Bay_area") %>%
  drop_na(rate) %>%
 #group_by(ID_type, time_period)%>%
  group_by(ID_type,time_period,region) %>%
  summarise(cumm_rate = sum (rate))

```
```{r, include=F}

my_new_bayarea_table <- my_bayarea_table %>%
  pivot_wider(names_from=c(ID_type),values_from= "cumm_rate")

```

\newpage

```{r}
# Table for  bay area only :
kable(my_new_bayarea_table, 
      booktabs=T, 
      col.names=c("Time_Period", " ", "Bacterial", "Fungal", "Parasitic", "Viral"),
      align='lccc', 
      caption="Infectious disease rates over time in the Bay Area from 2001-2018 by etiology of 
     disease and time period (3 year cummulatives)",
      format.args=list(big.mark=","))



```

\newpage



#Figures and codes: 
```{r, echo= F}
# Creating table for figure 1:
ID_tableyears_group_total_fig1Ltake2 <- ID_table3californiaonly %>%
 select(c("ID_type","rate_total","time_period")) %>%
   # time_period=as.character(time_period)%>%
  drop_na(rate_total) %>%
  group_by(time_period, ID_type)%>%
  mutate(average_rate=sum(rate_total)/3)%>%
  distinct(average_rate, .keep_all = TRUE)
 
ggplot(ID_tableyears_group_total_fig1Ltake2, aes(x = ID_type, y = average_rate)) +
geom_bar(aes(fill=time_period), stat="identity", position = position_dodge()) +
#geom_col(aes(fill=), col)
  scale_y_continuous(labels = function(x) format(x,bigmark=",",scientific=FALSE))+
  #scale_fill_manual(name= "time_period") +
  scale_fill_discrete(name= "time_period")+
  
  #values=c("#ffd333","#ff6600","#be0f24","#f91cc7","#910ff5","#003884")) +
labs(x="Group of Infectious disease", y = "Average 3 year rate",
title = "Figure 1: Trend of infectious diseases over time from 2001-2018 by type of 
     Disease and time period (3 year average of rate/100,000)")
```

Figure 1 shows that of the reported infectious diseases (excluding sexually transmitted diseases) that are most commonly reported are Bacterial diseases, followed by Fungal, and then parasitic diseases. Viral diseases have a lower 
rate. These numbers do not necessarily translates into real prevalence since 
many diseases are not considered "reportable", due to their common prevalence
and ubiquitous distribution. Ingeneral thorugh the years the frequency of
reported bacterial, Fungal and viral diseases have increased, while Parasitic
have decreased, except for 2016-2018 that shows an increasing trend. 




\newpage

```{r, echo =F}
#figure 2

#Figure 2L Trends from 2001-2018 only for Bacterial, Fungal and Parasitic conditions
ID_tableyears_group_total_fig2Ltake2 <- IDtable3 %>%
  select(c("ID_type","rate","year","cases",)) %>%
  mutate(year=as.character(year)) %>%
        # filter(region=="California")%>%
 filter(ID_type=="Bacterial"|ID_type== "Parasitic"|ID_type=="Fungal"|ID_type=="Viral") %>%
    #drop_na(rate) %>%
  group_by(ID_type, year)%>%
  summarize(sum_cases=sum(cases))%>%
  cbind(populationv)%>%
  rename(pop_total="...4")%>%
 mutate(rate_average= (sum_cases/pop_total)*100000)
 
#using plotly
plot_ly(
  ID_tableyears_group_total_fig2Ltake2,
  x= ~year,
  y= ~rate_average,
  color= ~ID_type,
  type="bar"
) %>%
  layout(barmode="stack")%>%
  
  layout(
    title = "Figure 2 : Trends of Bacterial Fungal and Parasitic Diseases
    rates per 100,000 from 2001-2018 in California
    (Excludes STD's)",
    xaxis = list(title = "Years"),
    yaxis = list(title = "California Rates per 100,000")
  )


```
Figure 2: This figure shows that since 2001, reports of bacterial diseases
have increased overtime. Reasons for this increase could be related to a real increase of reportable cases, versus improved report methodology. The same goes
to Fungal infections. Parasitic infections have decreased, except forthe period  2016-2018 that show an increase. Viral infection reports have increased since 2016 due to new viral reportable conditions like Zika and Chikungunya .



\newpage
```{r, echo=F}
#Take 2 figure 3 #Lourdes most common bacterial diseases from 2001 to 2018:
ID_mostcommon_bacterial_fig3Ltake2 <- IDtable3 %>%
  select(c("disease","rate","year","cases","population","sex","ID_type")) %>%
  mutate(year=as.character(year)) %>%
  filter(ID_type=="Bacterial") %>%
     drop_na(rate) %>%
 group_by(disease) %>%
  mutate(total_cases_average = sum(cases)/18) %>%
  distinct(total_cases_average, .keepall=TRUE)%>%
  filter(total_cases_average>120)
#%>%
  #summarize(cum_disease = sum(disease)) %>%
#mutate(average_cases= (cum_disease/18))

#using plotly
plot_ly(
  ID_mostcommon_bacterial_fig3Ltake2,
  x= ~disease,
  y= ~total_cases_average,
  color= ~disease,
  type="bar") %>%
# layout(barmode="stack")%>%
  
  layout(
    title = "Figure 3 :Most common reported Bacterial Infections
     average number of cases per year from 2001-2018 in California
    (Excludes STD's)",
    xaxis = list(title = "Bacterial Infections"),
    yaxis = list(title = "Average number of cases per year")
  )
```

Figure 3 : Among the bacterial infections, the most commonly reported one is Campilobacteriosis, followed by Salmonellosis and Shiguellosis.






```{r, echo =F}
#Take 2 Figure 4 will create a new bar chart for most common reported Parasitic diseases rates in California
ID_mostcommon_parasitic_fig4Ltake2 <- IDtable3 %>%
  select(c("disease","rate","year","cases","population","sex","ID_type")) %>%
  mutate(year=as.character(year)) %>%
  filter(ID_type=="Parasitic") %>%
     drop_na(rate) %>%
 group_by(disease) %>%
  mutate(total_cases_average = sum(cases)/18) %>%
  distinct(total_cases_average, .keepall=TRUE)%>%
  filter(total_cases_average>14)
#using plotly
plot_ly(
  ID_mostcommon_parasitic_fig4Ltake2,
  x= ~disease,
  y= ~total_cases_average,
  color= ~disease,
  type="bar") %>%
# layout(barmode="stack")%>%
  
  layout(
    title = "Figure 4 :Most common reported Parasitic Diseases
    yearly average 2001-2018 in California",
    xaxis = list(title = "Parasitic Infections"),
    yaxis = list(title = "Average number of cases per year")
  )
```
Figure 4: The most common parasitic disease is Giardiasis, followed by Amebiasis
and cryptosporidiosis.

```{r, include =F}
#Take 2 Figure 5  will create a new bar chart for most common reported fungal diseases rates in California
ID_mostcommon_fungal_fig5Ltake2 <- IDtable3 %>%
  select(c("disease","rate","year","cases","population","sex","ID_type")) %>%
  mutate(year=as.character(year)) %>%
  filter(ID_type=="Fungal") %>%
     drop_na(rate) %>%
 group_by( disease) %>%
mutate(total_cases_average = sum(cases)/18) %>%
  distinct(total_cases_average, .keepall=TRUE)

#using plotly
plot_ly(
  ID_mostcommon_fungal_fig5Ltake2,
  x= ~disease,
   y= ~total_cases_average,
   color= ~disease,
  type="bar")%>%
# layout(barmode="stack")%>%
  
  layout(
    title = "Figure 5 : Most common reported Fungal Infections
    from 2001-2018 in California
    (Excludes STD's)",
    xaxis = list(title = "Fungal infections"),
    yaxis = list(title = "average numeber of cases per year")
  )
```



\newpage
```{r, echo=F}
#take 2 Lourdes; will create a new bar chart for most common reported viral 
#diseases in California
ID_mostcommon_viral_fig6Ltake2 <- IDtable3 %>%
  select(c("disease","rate","year","cases","population","sex","ID_type")) %>%
  mutate(year=as.character(year)) %>%
  filter(ID_type=="Viral") %>%
     drop_na(rate) %>%
 group_by(disease) %>%
  mutate(total_cases = sum(cases)) %>%
  distinct(total_cases, .keepall=TRUE)
  #filter(total_cases>14)
  
  
#using plotly
plot_ly(
  ID_mostcommon_viral_fig6Ltake2,
  x= ~disease,
  y= ~total_cases,
  color= ~disease,
  type="bar")%>%
# layout(barmode="stack")%>%
  
  layout(
    title = "Figure 5 :Most common reported Viral Infections
    total cases from 2001-2018 in California
    (Excludes STD's)",
    xaxis = list(title = "Viral infections"),
    yaxis = list(title = "Total cases")
  )
```
Figure 5 : Among viral infections, the most commonly reported was Dengue virus infection. The newly described virus Chikungunya and Zika virus were not
reported in California until 2017




\newpage

##Analyzing STDs:

#Methods:
The visualiztion that I would like to create is a graph rates of each disease type (bacteria, virus, std etc) overtime within the Bay Area. To do this, I first filtered both the group dataset and the STD dataset to only Bay Area Counties ("Alameda", "Santa Clara", "San Mateo", "San Francisco", "Marin", "Contra Costa", "Solano")

```{r, include=F}

#sandya-create disease trend over time

individualdata1<- read_csv("stds-by-disease-county-year-sex.csv")
groupdata<-read_csv("idb_odp_2001-2018 (1) (1).csv")

groupdatafinal<-filter(groupdata, County %in% c("ALAMEDA", "SANTA CLARA", "SAN MATEO", "SAN FRANCISCO", "MARIN", "CONTRA COSTA", "SOLANO"))


  
individualdatafinal<-filter(individualdata1, County %in% c("Alameda", "Santa Clara", "San Mateo", "San Francisco", "Marin", "Contra Costa", "Solano"))%>% select(-10)


combineddata<-rbind(groupdatafinal, individualdatafinal)

combineddata$County<-tolower(combineddata$County)
combineddata$Sex<-tolower(combineddata$Sex)

combineddata1<-combineddata%>%filter(Sex %in% "total")

combinedatafinal<-combineddata1%>%mutate(Disease_Type = case_when(Disease %in% c("Gonorrhea", "Early Syphilis", "Chlamydia")~ "STD (Bacterial)", Disease %in% c("Shiga toxin-producing E. coli (STEC) with Hemolytic Uremic Syndrome (HUS)",      "Shiga toxin-producing E. coli (STEC) without Hemolytic Uremic Syndrome (HUS)", "Anaplasmosis and Ehrlichiosis", "Hemolytic Uremic Syndrome (HUS) without evidence of Shiga toxin-producing E. coli (STEC)", "Paratyphoid Fever", "Ehrlichiosis", "Anaplasmosis" ,"Shiga Toxin Positive Feces (without culture confirmation)", "E. coli Other STEC (non-O157)",  "Yersiniosis"  ,"Vibrio Infection (non-Cholera)" , "Typhoid Fever", "Typhus Fever",  "Tularemia"  ,"Streptococcal Infection (cases in food and dairy workers)"  ,"Spotted Fever Rickettsiosis" , "Shigellosis","Salmonellosis", "Relapsing Fever" , "Q Fever" , "Psittacosis" , "Plague, human","Lyme Disease", "Listeriosis"  ,"Leptospirosis","Leprosy (Hansen's Disease)" ,"Legionellosis","E. coli O157" ,"Cholera"  , "Campylobacteriosis"  ,"Brucellosis" , "Anthrax" )~ "Bacteria", Disease %in% c("Chikungunya Virus Infection", "Dengue Virus Infection","Flavivirus Infection of Undetermined Species", "Hantavirus Infection","Hepatitis E, acute infection", "acute infection", "Rabies, human","Yellow Fever", "Zika Virus Infection")~ "Virus", Disease %in% c("Amebiasis","Babesiosis", "Cryptosporidiosis", "Cyclosporiasis", "Cysticercosis or Taeniasis", "Malaria", "Giardiasis", "Trichinosis")~"Protozoa", Disease %in% c( 'Botulism, Foodborne', 'Botulism, Other', 'Botulism, Wound', 'Ciguatera Fish Poisoning', 'Domoic Acid Poisoning',  'Paralytic Shellfish Poisoning', 'Scombroid Fish Poisoning')~ "Toxin", Disease %in% c('Creutzfeldt-Jakob Disease and other Transmissible Spongiform Encephalopathies')~ "Prion", Disease ==  'Hemolytic Uremic Syndrome (HUS)'~ "Infectious Complication", Disease == "Coccidioidomycosis"~ "Fungal"))

```
```{r, include=F}
data1<-combinedatafinal%>%select(-9)



plotdatatest<-data1%>%group_by(Disease_Type, Year)%>%summarize(Sum_cases=sum(Cases))


testin<-data1%>%group_by(County, Year)%>%summarize(total_pop=mean(Population))%>%group_by(Year)%>%summarise(totalp=sum(total_pop))

hope<-left_join(testin, plotdatatest, by= "Year")
hope$Overall_Rate<-(hope$Sum_cases/hope$totalp)*100000


```


\newpage
```{r}
ggplot(hope, aes(x=Year, y=Overall_Rate))+facet_wrap(vars(Disease_Type), ncol = 2, scales = "free_y")+geom_line(aes(color = Disease_Type))+labs(x="Year", y="Overall Rate (per 100K)", title = "Figure 7: Overall Rates of Infectious diseases (including STDs) in Bay Area Counties from 2001-2018") + theme_minimal()
```
Interpretation of graph: This graph looks at the overall rates per year of 
different types of infectious disease from 2001 to 2018 in the Bay area counties, with the Y axis adjusted. I created this graph to better visualize the trends. From the graphs it is noticeable that fungal and 
bacterial rates are increasing overtime, along with STD rates. 

\newpage
```{r, include=F}
combineddatagender<-individualdatafinal%>%filter(Sex %in% c("Male", "Female"))%>%select(-9)%>%group_by(Sex, Year)%>%summarise(case_total=sum(Cases))
 
std_set<-left_join(testin, combineddatagender, by= "Year")
std_set$Overall_Rate<-(std_set$case_total/std_set$totalp)*100000

```

```{r}
ggplot(std_set, aes(x=Year, y=Overall_Rate))+geom_line(aes(color = Sex))+labs(x="Year", y="Overall Rate (per 100K)", title = "Figure 8: Overall rates per year of STDs (Bacterial) in Bay Area Counties from 2001-2018 seperated by Sex") +theme_minimal()


```
Interpretation of graph: This graph looks at the Overall rates per year of 
bacterial STD infectious disease in the Bay area counties from 2001 to 2018, seperated by Sex. I created this graph to better visualize the trends in STDs 
between males and females.The graphs shows a very significant increase in the 
overall rate of STDs for both male and females. Prior to around 2014, it seems
that female rates were higher than male rates. However from around 2014 and 
onward, we see an even greater increase in male rates.

```{r}
tabledataSTD<-individualdatafinal%>%filter(Sex %in% "Total")%>%group_by(Disease, Year)%>%summarise(STD_case_total=sum(Cases))
```

```{r}

newSTD<-left_join(testin, tabledataSTD, by= "Year")
newSTD$Overall_Rate<-(newSTD$STD_case_total/newSTD$totalp)*100000
newSTD1<-newSTD%>%select(1,3,5)%>% pivot_wider(names_from="Year",values_from= "Overall_Rate")%>%select(1, 15,16,17,18,19)
formattable(newSTD1)

```

Interpretation of table: This table is a visualization of the rates of STDs in the Bay Area over the last 5 years that were included in the dataset (2014-2018). As shown in the table, rates of all three STDs are increasing significantly each year. 